---
title: "A Statistician's Priority List for Boosting Home Value"
theme: simplex
output:
  word_document: default
  html_document: default
---

## Introduction

For many people, a house is not only a residence but also a place where they have been investing throughout their stay. Hence, how to maintain and boost home value during their stay has been a question for many house owners. Generally speaking, there are many commonly known factors that would help increase home values, however, for most people with a limited budget, it is hard to take everything into consideration when they want to boost their home values. Therefore, it is of great significance to learn what should be prioritized during the home improvement with a purpose of value boost. Although home onwers are unable to obtain everything they want with a tight budget, they can do the things that really matter and bring up the sale prices of the house by prioritizing the controllable things. To obtain a priority list for house improvement and home value bringing up, statistical methods like linear regression and random forest would be utilized in this project to analyze a Kaggle dataset containing house sale prices of King County, Washington from May 2014 to May 2015. Statistical models would be constructed to find out the most significant house attributes that are related to home prices. 

## Purpose

The primary objective of this project is to find out a priority list for home improvement that can be helpful for boosting home values from a statistician's view. Commonly used house attributes would be analyzed and the project would be planned to figure out the most relevant features of a house regarding sale prices. Hopefully, this project could offer some suggestions on house improvement and home value boosting for home investors during their stay.

## Data

A real-world dataset that contains house sale price information and the corresponding house features of King County, Washington from 2014 to 2015 will be used. It is originated from Kaggle, and can be imported to R from `mlr3data` package. 

Basically, there are 21,613 observations along with 19 house features such as the number of bathrooms, bedrooms, floors, and square footage of the housein the original data. The code that help us load the data and print the fisrt few lines of the original data is shown as below

```{r}
## Load the required dataset
library(mlr3data)
data("kc_housing")
head(kc_housing)
```

```{r}
## Print out the summary statistics
summary(kc_housing)
```

```{r}
## Print out the data dimensions
dim(kc_housing)
```

## Variables

The full variable dictionary is summarized as below:

**id:** unique ID of the house

**date:** the sale date of the house

**price:** the final sale price of the house

**bedrooms:** count of bedrooms in the house

**bathrooms:** count of bathrooms in the house

**sqft_living:** square footage of the living area in the house

**sqft_lot:** square footage of the lot for the house

**floors:** total levels in the house

**waterfront:** whether the house has a waterfront view. If yes, the value is 1. Otherwise, the value is 0.

**view:** how many times the house has been viewed

**condition:** the overall condition of the house

**grade:** the overall grade given to the housing unit by King County grading system. According to [King County Assessor's webpage](http://info.kingcounty.gov/assessor/esales/Glossary.aspx?type=r), this represents the construction quality of improvements. Grades run from grade 1 to 13.

**sqft_basement:** square footage of the basement

**sqft_above:** square footage of the house apart from the basement

**yr_built:** which year the house was built

**yr_renovated:** which year the house was renovated. If no renovation has been done, the value is 0

**zipcode:** the zip code for the house address

**lat:** latitude coordinate of the house location

**long:** longitude coordinate of the house location

**sqft_living15:** square footage of the living area in the house measured in 2015

**sqft_lot15:** square footage of the lot for the house measured in 2015

**renovated:** whether the house has been renovated. If yes, the value is 1. Otherwise, the value is 0.

**basemt:** whether the house has basement. If yes, the value is 1. Otherwise, the value is 0.


## Packages

The first package that will be used in this project is `mlr3data`, which offers the dataset that we are going to analyze. Besides, we will use the `ggplot2` and `lattice` packages for the purpose of data visualization and `randomForest` for random forest models that can be helpful to analyze the effects of the house factors on the house price. Also, we will utilize the `stargazer` package to offer neat and more readable model results of linear regressions. Anotehr important package that can be useful in this project is `GGally`, in which the `ggcorr` function can help us obtain the correlation matrix. Finally, we also use the `dplyr` package to manipulate and modify data frames.

```{r message=FALSE, warning=FALSE}
## Load the required packages
library(ggplot2)
library(lattice)
library(randomForest)
library(stargazer)
library(GGally)
library(dplyr)
```

## Exploratory Data Analysis


We start our data exploration with the variable of interest `price`.

```{r EDA - Price, fig.align = "center"}
# Check the distribution of house sale price
qplot(x = price, data = kc_housing, bins = 50,
      main = "King County House Sale Prices")
```

```{r summary}
# 5-point summary of price
summary(kc_housing$price)
```

It is clear that the distribution of the home price is positively skewed with a long right tail, which implies that some houses are expected to have higehr values than others.

Next, we explore the relationship between the features of the house and the home prices by plotting `price` with each feature. We find that some features like  `bathrooms`, `sqft_living`, `grade` and `waterfront` have relatively stronger relationships with `price` than others. 

```{r graphs, message=FALSE, warning=FALSE}
# Create scatterplot for price and bathrooms
ggplot(kc_housing, aes(x = bathrooms, y = price)) +
  geom_point(shape = 1) +
  geom_smooth(method = lm, color = "orange", se = FALSE) +
  ggtitle("Scatterplot of Price vs. Number of Bathrooms")
# Create scatterplot for price and sqft_living
ggplot(kc_housing, aes(x = sqft_living, y = price)) +
  geom_point(shape = 1) +
  geom_smooth(method = lm, color = "orange", se = FALSE) +
  ggtitle("Scatterplot of Price vs. Square Footage of Living Area")
# Create scatterplot for price and grade
ggplot(kc_housing, aes(x = grade, y = price)) +
  geom_point(shape = 1) +
  geom_smooth(method = lm, color = "orange", se = FALSE) +
  ggtitle("Scatterplot of Price vs. Grade")
# Create scatterplot for price and waterfront
ggplot(kc_housing, aes(x = waterfront, y = price)) +
  geom_point(shape = 1) +
  geom_smooth(method = lm, color = "orange", se = FALSE) +
  ggtitle("Scatterplot of Price vs. Waterfront")
```

We also include the correlation matrix that reflects how variables are correlated with each other.

```{r correlation matrix, fig.height = 10, fig.width = 15, message = FALSE, warning = FALSE, fig.align = "center"}
# Correlation matrix for numeric variables
ggcorr(kc_housing[, -c(1:2)], method = c("everything","pearson"))
```

We noticed that `sqft_living` and `sqft_above` are highly correlated with a correlation of `r cor(kc_housing$sqft_living, kc_housing$sqft_above)`. This makes a lot of sense because most of living area is usually above the basement. The univariate correlation between `sqft_living` and `price` (`r cor(kc_housing$sqft_living, kc_housing$price)`) is higher than that between `sqft_above` and `price` (`r cor(kc_housing$sqft_above, kc_housing$price)`). Similarly, `sqft_living` and `sqft_living15` are highly correlated with a correlation of `r cor(kc_housing$sqft_living, kc_housing$sqft_living15)`. The univariate correlation between `sqft_living` and `price` (`r cor(kc_housing$sqft_living, kc_housing$price)`) is higher than that between `sqft_living15` and `price` (`r cor(kc_housing$sqft_living15, kc_housing$price)`). 

## Data Modification

Based on above findings, we will modify our dataset by introducing two new binary features:

- renovated: Equal to 1 if the house have been renovated and 0 otherwise

- basemt: Equal to 1 if a house has basement and 0 otherwise. 

We also exclude the useless information from the original set and remain the variables that are most relevant to the house price. Moreover, we would drop the variables of sqft_above and sqft_living15 in the further analysis to avoid the issue of multicollinearity. The new modified version of the dataset is named as house, and it’s summary statistics are printed.

```{r Create a new dataset}
# Create a new dataset for further analysis
# Create new variable "renovated" based on the existing variable "yr_renovated"
kc_housing$renovated = as.factor(ifelse(kc_housing$yr_renovated > 0, "1", "0"))
# Create new variable "basemt" based on the existing variable "sqft_basement"
kc_housing$basemt = as.factor(ifelse(kc_housing$sqft_basement > 0, "1", "0"))
house = subset(
  kc_housing,
  select = -c(
    date,
    sqft_basement,
    sqft_living15,
    sqft_above,
    sqft_lot,
    sqft_lot15,
    yr_built,
    yr_renovated,
    zipcode
  )
)
house = na.omit(house)
summary(house)
```

## Modeling

- Random Forest Model

```{r message=FALSE, warning=FALSE}
## Fit the best random forest model
rf_mod = randomForest(price ~ .,
                      data = house,
                      mtry = 7,
                      importance = T)
importance = importance(rf_mod)
VarImportance = data.frame(variables = row.names(importance),
                           importance = round(importance[, '%IncMSE'], 2))

## Rank variables by importance
rank = VarImportance %>% mutate(rank = paste0('#', dense_rank(desc(importance))))

ggplot(rank, aes(
  x = reorder(variables, importance),
  y = importance,
  fill = importance
)) +
  geom_bar(stat = 'identity') +
  geom_text(
    aes(x = variables, y = 0.5, label = rank),
    hjust = 0,
    vjust = 0.6,
    size = 4,
    color = 'white'
  ) +
  labs(x = 'Variables', y = 'Importance') +
  ggtitle("Variable Importance from Selected Random Forest Model") +
  coord_flip()
```

It turns out that the `grade` (ranging from 1 to 13), which represents the construction quality, is the most important variable regarding house sale price. For houses only meet the minimum building standards, their grades are low and ranged from 1 to 3, and their prices are expected to be the lowest on average. For houses that have achieved average performance in terms of construction and design, they are graded as 7. And their averaged prices are expected to be moderate among all houses. As for the houses graded over 12, they are thought to have excellent designs and use the best materials while construction. As a result, their prices are expected to be highest too. 

`sqft_living`, the spacing of the living rooms, turns out to be the next most important feature that is related to home values. And the further next significant factor that is highly correlated with the home prices is the location, consistent with the `latitude` and `longitude` ranked as third and fourth important variables. 

- Multiple Linear Regression Model

Although, we learned the effects of location on house price are likely to be significant, we are not going to interpret them in the multiple linear regression model because location is not a factor that could be changed for home owners after the house being purchased. Hence, in this linear model, only the variables of `grade`, `sqft_living`, `view`, `waterfront` and `bathrooms` will be included.

```{r  message=FALSE, warning=FALSE}
## Fit the linear regression model
lm.mod = lm(log(price) ~ grade + sqft_living + view + bathrooms +
              waterfront,
            house)
lm.mod = step(lm.mod, trace = FALSE)
## Print the model results
stargazer(lm.mod, type = "text")
```

The R-squared of this model is 0.711, which means this linear model containing the most relevant predictors could explain about 71.1% of the total variation in the house prices. And F statistic of this model is 0.317 (p = 0.000), which means the overall model is statistically significant. Also, at 10% level of significance, we notice all predictors are statistically significant individually. 

As we notice, here we used the log-transformation on the dependent variable because the variable is right skewed based on previous analysis. Hence, the model results imply that one additional point in the house grade is expected to increase the house price by 25.1% on average when other factors are assumed to be the same. And the house with a waterfront is expected to be 18.4% higher in price than the house without a waterfront when other conditions are the same. And an additional bathroom in a house is expected to bring up the home value by 4.3%, and with a one-sqft increase in the living room is expected to enhanced the house value by 0.01% on average conditional to all other factors respectively. As for the time of the house being viewed, it is shown that each time the house is viewed, the house value is expected to be boost by 4.4% on average.




